---
title: The Ground Game
author: Hannah Valencia
date: '2022-10-16'
slug: []
categories: []
tags: []
summary: "This week we learned about how election turnout can affect the outcome of races, and how on-the-ground campaigning efforts aim to mobilize and persuade voters. We can look at this data on the district-level to uncover patterns, and compare it to last week's data on ad spending to see how they interact."
---

*This blog is part of a series related to Gov 1347: Election Analytics, a course at [Harvard University](https://www.harvard.edu/) taught by Professor [Ryan D. Enos](http://ryandenos.com/)*.

*This week we learned about how election turnout can affect the outcome of races, and how on-the-ground campaigning efforts aim to mobilize and persuade voters. We can look at this data on the district-level to uncover patterns, and compare it to last week's data on ad spending to see how they interact.*


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# load libraries

library(tidyverse)
library(usmap)
library(plotly)
library(rmapshaper)
library(sf)
library(haven)
library(sjlabelled)
library(usdata)
library(stargazer)

```


```{r}
poll <- read_csv("week3_data.csv")
house <- read_csv("house party vote share by district 1948-2020.csv")
expert_rating <- read_csv("expert_rating.csv")
house_polls_long <- read_csv("house_polls_long.csv")
ratings2018 <- read_csv("2018_ratings_share.csv")
cvap_2012_2020 <- read_csv("cvap_district_2012-2020_clean.csv")
dist_polls <- read_csv("dist_polls_2018-2022.csv")
```































