---
title: Economic Forces
author: Hannah Valencia
date: '2022-09-19'
slug: []
categories: []
tags: []
summary: "This week, we looked at using various economic variables as predictors for election outcomes. In this blog, I aim to predict Republican seat share and incumbent party vote share using variables such as GDP growth percentage, CPI percent change, disposable income percent change, and the national unemployment rate."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(car)
library(ggplot2)
library(sf)
library(readr)
library(jtools)
library(lfe)
library(corrplot)
```

<br>

*This week, we looked at using various economic variables as predictors for election outcomes. In this blog, I aim to predict Republican seat share and incumbent party vote share using variables such as GDP growth percentage, CPI percent change, disposable income percent change, and the national unemployment rate.*

<br>

```{r, warning=FALSE, message=FALSE}

# load data
popvote_df <- read_csv('election-section-2/house_popvote_seats.csv') # house and seat share by year
economy_df <- read_csv('election-section-2/GDP_quarterly.csv') # GDP data by quarter
state_df <- read_csv('election-section-2/unemployment_state_monthly.csv') # state unemployment data
rdi_df <- read_csv('election-section-2/RDI_quarterly.csv') # rdi
cpi_df <- read_csv('election-section-2/CPI_monthly.csv') #cpi 
unemp_df <- read_csv('election-section-2/unemployment_national_quarterly_final.csv') # national unemploymend data

```

```{r, warning=FALSE, message=FALSE}

# Clean Data
# merge
dat <- left_join(economy_df, popvote_df, by = 'year')

# drop NAs (only election years)
dat <- dat %>%
  drop_na()

# new df
dat2 <- dat %>% 
    select('year', 'winner_party', 'H_incumbent_party', 'H_incumbent_party_majorvote_pct', 
           'quarter_cycle', 'GDP_growth_pct') %>%
    filter(quarter_cycle == 7)  

# merge    
data <- left_join(rdi_df, popvote_df, by = 'year')

# drop NAs (only election years)
data <- data %>%
  drop_na()

# new df
data2 <- data %>% 
    select('year', 'winner_party', 'H_incumbent_party', 'H_incumbent_party_majorvote_pct', 
           'quarter_cycle', 'DSPIC_change_pct') %>%
    filter(quarter_cycle == 7) 

# using quarter_cycle 3 instead of 7 because the years in this dataframe do not align with the years in the other dataframes. Here, quarters 1-4 are in 1948 while in the other dataframes, quarters 5-8 are in 1948. 3 is the quarter that aligns with 7 in terms of the time of year (July - August).
data3 <- unemp_df %>%
  select('year', 'UNRATE', 'quarter_cycle') %>%
  filter(quarter_cycle == 3) 

# merge data2 and dat2
data2$GDP_growth_pct <- dat2$GDP_growth_pct[match(data2$year, dat2$year)]
data2$unrate <- data3$UNRATE[match(data2$year, data3$year)]

```

```{r, warning=FALSE, message=FALSE}

# create month and year cols for CPI
cpi_df$year <- as.numeric(format(as.Date(cpi_df$DATE, format="%Y-%m-%d"),"%Y"))
cpi_df$month <- as.numeric(format(as.Date(cpi_df$DATE, format="%Y-%m-%d"),"%m"))

# create quarter variable for CPI - only 4 to make it easier but can use year to change it into 8
# create a CPI quarterly variable that averages the CPI across the months of each quarter
# create a CPI change variable that shows CPI change from the previous quarter since people are likely to feel CPI relatively
cpi_df <- cpi_df %>%
  mutate(quarter = case_when(month == c(1:3) ~ "1",
                             month == c(4:6) ~ "2",
                             month == c(7:9) ~ "3",
                             TRUE ~ "4")) %>%
  group_by(year, quarter) %>%
  mutate(CPI_quarter_avg = mean(CPIAUCSL)) %>%
  ungroup() %>%
  mutate(CPI_pct_change = (CPI_quarter_avg/lag(CPI_quarter_avg, n = 3) - 1) * 100) %>%
  filter(quarter == 3)

data2$CPI_pct_change <- cpi_df$CPI_pct_change[match(data2$year, cpi_df$year)]

```

So far, we have touched on a few different forces that affect the way in which people vote -- aside from simply party affiliation. Last week, we learned about end heuristics, or how voters tend to focus on the period of time right before an election rather than an incumbent's entire time in office. If the candidate ends their term strong, that is what people are likely to remember. I have brought that idea into this week's analysis of economic forces -- choosing to focus on Quarter 7, or July through August in an election year.
<br>

According to Lewis-Beck and Stegmaier (2000), voters tend to weigh economic issues more heavily than other issues, and they "hold the government responsible for economic performance". While the government can only do so much to mitigate the economy's natural cycles, people still hold those in office responsible for it all. It should also be noted that Republicans give more weight to economic issues than Democratic voters. The economic variables that I will consider in this analysis are GDP growth percentage, percent change in disposable income, percent change in Consumer Price Index (CPI), and the national unemployment rate.
<br>

In order to make predictions, I have created four different linear regression models that build off of one another in the variables used to predict the incumbent party's vote share percent. The first starts off with only GDP growth percent as an independent variable, and we find this is not a good model for prediction. As shown in the regression table under "Model 1", the coefficient on GDP growth is insignificant and the model has an R-squared value of 0.00, telling us that 0% of the variation in incumbent party vote share can be explained by GDP growth percent in this model. This model does not fit the data well at all.
<br>

We expand on this model by then adding an independent variable for the percent change in disposable income (Model 2), and then, in a different model, another independent variable for the unemployment rate (Model 3). In both of these models, all the independent variables' coefficients remain insignificant, although we do see a *slight* rise in the R-squared value. In Model 2, the R-square value is 0.02, and in Model 3 it is 0.04. While this shows a slight increase from Model 1, these numbers are still indicative of a model that does not fit the data well at all.
<br>

In the final model, I added an independent variable for the percent change in CPI. Here (Model 4), this CPI variable is significant at the 5% level. For every 1% change in CPI, the incumbent party vote share percent increases by 1.55 on average. In a healthy economy, some inflation is expected, so this increase in CPI leading to an increase in the incumbent's vote share percent could indicate that people see rising prices as a positive thing. This model also has an R-squared value of 0.23, showing that it is a much better fit for the data than the previous three models. I coded the CPI percent change variable by changing the monthly CPI statistic into a quarterly average, and then created the CPI change variable by lagging the CPI from the previous quarter and converting it into a percent change with respect to that quarter's CPI. I believed this would give an accurate representation of inflation in the economy, and is more understandable than the CPI index number that shows the price of a specific basket of goods. 
<br>

While the relationship between CPI percent change and vote share is positive in this model shows, current events could change this relationship, or at least create an inflation threshold where it goes from being seen as a good thing to a bad thing. Our inflation levels are so high right now that the CPI index is likely negatively correlated with how the incumbent party will do in the upcoming election.
<br>

To look at how these models fit over time, I created a graph that shows each model's predictions alongside the actual values of incumbent party vote share. Each model is a colored, dotted line, while the actual values are represented by the black, solid line. As we can see, all of the colored lines do not have as much variation as the true values line; they are all much flatter. Only the green line representing Model 4 comes close to following the patterns of the actual vote share values, but even it cannot recreate the variation we see over time. This model that used CPI did a decent job of predicting between 1970 and 2000. After 2000, however, it did not hit the lows that we saw with the actual values. Until the 1990s, incumbent vote share typically sat above 52%. After this period and into the 2000s, however, the incumbent was not as successful, often falling into the mid 40-percents. It is during this time period that the residuals appear to grow.
<br>


```{r, results='asis'}

# national-level reg
econ1 <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct, 
              data = data2)
econ2 <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct + DSPIC_change_pct, 
              data = data2)
econ3 <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct + DSPIC_change_pct + unrate, 
              data = data2)
econ4 <- lm(H_incumbent_party_majorvote_pct ~ GDP_growth_pct + DSPIC_change_pct + unrate + CPI_pct_change,
              data = data2)

export_summs(econ1, econ2, econ3, econ4, scale=TRUE,
             coefs = c("Intercept"= "(Intercept)",
                       "GDP Growth Percentage" = "GDP_growth_pct",
                       "Disposable Income Percent Change" = "DSPIC_change_pct",
                       "Unemployment Rate" = "unrate",
                       "CPI Percent Change" = "CPI_pct_change"))


# use model predictors on vote share
plot(data2$year, data2$H_incumbent_party_majorvote_pct, 
     type="l",
     main="True and Predicted Vote Share by Year",
     xlab = "Year",
     ylab = "Incumbent Party Vote Share")
lines(data2$year, predict(econ4, data2), col = "green", lty = 3)
lines(data2$year, predict(econ3, data2), col = "blue", lty = 3)
lines(data2$year, predict(econ2, data2), col = "red", lty = 3)
lines(data2$year, predict(econ1, data2), col = "orange", lty = 3)

legend("bottomleft", c("Actual VS", "Model 4 Prediction", "Model 3 Prediction", "Model 2 Prediction", "Model 1 Prediction"),
       lty = c(1, 3, 3, 3, 3, 3),
       col = c("black", "green", "blue", "red", "orange"))

```
<br>
I repeated this modelling process and prediction but instead using Republican seat share percentage as the dependent variable. The first three models that do not use CPI percent change again came up poorly fitting models for the data, with no significant independent variables. Only once percent change of CPI is added does the model have a relevant r-squared value of 0.30, and the CPI dependent variable is significant at the 1% level. These models demonstrate the same effects for incumbent vote share as for Republican seat share.
<br>

Once again, I created a plot of the predictions versus the true values over time. Interestingly, we see the same pattern of the CPI model fairly accurately fitting the actual values until the 1990s, after which they diverge and the residuals grow. The other three models appear as flat lines in this graph, similar to the plot above.
<br>

```{r, warning = FALSE, message = FALSE}

# repeat analysis for seat share
popvote_df <- popvote_df %>%
  mutate(RepSeatShare = R_seats/(R_seats + D_seats + Other_seats),
         DemSeatShare = D_seats/(R_seats + D_seats + Other_seats))

# add new seat share vars to df with other econ data
data2$RepSeatShare <- popvote_df$RepSeatShare[match(data2$year, popvote_df$year)]
data2$DemSeatShare <- popvote_df$DemSeatShare[match(data2$year, popvote_df$year)]

```

```{r, results='asis'}

ss1 <- lm(RepSeatShare ~ GDP_growth_pct, 
              data = data2)
ss2 <- lm(RepSeatShare ~ GDP_growth_pct + DSPIC_change_pct, 
              data = data2)
ss3 <- lm(RepSeatShare ~ GDP_growth_pct + DSPIC_change_pct + unrate, 
              data = data2)
ss4 <- lm(RepSeatShare ~ GDP_growth_pct + DSPIC_change_pct + unrate + CPI_pct_change,
              data = data2)

export_summs(ss1, ss2, ss3, ss4, scale=TRUE,
             coefs = c("Intercept"= "(Intercept)",
                       "GDP Growth Percentage" = "GDP_growth_pct",
                       "Disposable Income Percent Change" = "DSPIC_change_pct",
                       "Unemployment Rate" = "unrate",
                       "CPI Percent Change" = "CPI_pct_change"))

# use model predictors on seat share
plot(data2$year, data2$RepSeatShare, 
     type="l",
     main="True and Predicted Seat Share by Year",
     xlab = "Year",
     ylab = "Republican Party Seat Share")
lines(data2$year, predict(ss1, data2), col = "green", lty = 3)
lines(data2$year, predict(ss2, data2), col = "blue", lty = 3)
lines(data2$year, predict(ss3, data2), col = "red", lty = 3)
lines(data2$year, predict(ss4, data2), col = "orange", lty = 3)

legend("topleft", c("Actual Seat Share", "Model 1 Pred", "Model 2 Pred", "Model 3 Pred", "Model 4 Pred"),
       lty = c(1, 3, 3, 3, 3, 3),
       col = c("black", "green", "blue", "red", "orange"))

```

<br>
One issue to note when analyzing the effects of different economic forces jointly on a dependent variable is multicollinearity. This is a problem that occurs when multiple independent variables are correlated with one another and negatively impacts the reliability of the inferences we can draw from these models. To assess if multicollinearity exists between any of the independent variables, I have created a correlation matrix below.

<br>
```{r}
corrmat <- data2[, c(4, 6:9)]
res <- cor(corrmat)
corrplot(res, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45, addCoef.col = "grey70")
```
<br>
We can see that GDP growth percentage is highly negatively correlated with the percent change in disposable income, with a correlation of -0.72. There is also a moderately strong correlation between the unemployment rate and the percent change of disposable income. Both of these correlations may impact the validity of our models and hurt any insights we draw from this analysis.








