---
title: The Ground Game
author: Hannah Valencia
date: '2022-10-16'
slug: []
categories: []
tags: []
summary: "This week we learned about how election turnout can affect the outcome of races, and how on-the-ground campaigning efforts aim to mobilize and persuade voters. We can look at this data on the district-level to uncover patterns, and compare it to last week's data on ad spending to see how they interact."
---

*This blog is part of a series related to Gov 1347: Election Analytics, a course at [Harvard University](https://www.harvard.edu/) taught by Professor [Ryan D. Enos](http://ryandenos.com/)*.

*This week we learned about how election turnout can affect the outcome of races, and how on-the-ground campaigning efforts aim to mobilize and persuade voters. We can look at this data on the district-level to uncover patterns, and compare it to last week's data on ad spending to see how they interact.*


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

# load libraries

library(tidyverse)
library(usmap)
library(plotly)
library(rmapshaper)
library(sf)
library(haven)
library(stringr)
library(sjlabelled)
library(usdata)
library(stargazer)

```


```{r}
poll <- read_csv("week3_data.csv")
generic_polls <- read_csv("GenericPolls1942_2020.csv")
house <- read_csv("house party vote share by district 1948-2020.csv")
expert_rating <- read_csv("expert_rating.csv")
house_polls_long <- read_csv("house_polls_long.csv")
ratings2018 <- read_csv("2018_ratings_share.csv")
cvap_2012_2020 <- read_csv("cvap_district_2012-2020_clean.csv")[, -c(1:2)]
dist_polls <- read_csv("dist_polls_2018-2022.csv")
fund1 <- read_csv("data1.csv")[, -c(1:2)]
econ <- read_csv("data2.csv", skip = 1)[, -1]
state_econ <- read_csv("state_df.csv")[, -1]
incumb_dist <- read_csv("incumb_dist_1948-2020 (3).csv")[, -c(1:2)]
```

```{r}

# FIXING DATA FROM OTHER WEEKS TO ADD TO MASTER

# mutating the state-level economic variables to get a Q3 mean unemployment rate for each state from 2014-2021

state_econ1 <- state_econ %>%
  filter(Year >= 2014,
         Month %in% c(7:9)) %>%
  select(`FIPS Code`, `State and area`, Month, Year, Unemployed_prct) %>%
  group_by(`State and area`, Year) %>%
  pivot_wider(names_from = Month, values_from = Unemployed_prct) %>%
  mutate(Q3_unemployed = sum(`7`, `8`, `9`)/3)


# mutating the national economic data for House incumbent and House winner from 2014-2020

econ1 <- econ %>%
  filter(year >= 2014) %>%
  select(year, winner_party, H_incumbent_party) %>%
  rename(H_winner_party = winner_party)


# mutating the pollster expert ratings at the district level for average poll rankings on the 1-7 scale

expert_rating1 <- expert_rating %>%
  filter(year >= 2014) %>%
  select(year, state, district, cook, rothenberg, sabatos_crystal_ball) %>%
  group_by(year, state, district) %>%
  mutate(avg_rating = sum(cook, rothenberg, sabatos_crystal_ball)/3) %>%
  select(year, state, district, avg_rating)

# formatting the district to match the master df
expert_rating1$district[expert_rating1$district == "AL"] <- "00"
expert_rating1$district[expert_rating1$district == "1"] <- "01"
expert_rating1$district[expert_rating1$district == "2"] <- "02"
expert_rating1$district[expert_rating1$district == "3"] <- "03"
expert_rating1$district[expert_rating1$district == "4"] <- "04"
expert_rating1$district[expert_rating1$district == "5"] <- "05"
expert_rating1$district[expert_rating1$district == "6"] <- "06"
expert_rating1$district[expert_rating1$district == "7"] <- "07"
expert_rating1$district[expert_rating1$district == "8"] <- "08"
expert_rating1$district[expert_rating1$district == "9"] <- "09"

# adding 2018 ratings

ratings2018_1 <- ratings2018 %>%
  select(District, avg) %>%
  separate(District, c("state_abb", "district"), "-")

# changing the state from abbreviations to names
ratings2018_1$state <- state.name[match(ratings2018_1$state_abb, state.abb)]
ratings2018_1$year <- 2018

ratings2018_1 <- ratings2018_1 %>%
  select(year, state, district, avg) %>%
  rename(avg_rating_2018 = avg)


# mutating generic ballot data - want only polls conducted in the 30 days until the election -> average those polls
# gives us the average support for the party in the month leading up to the election

gp1 <- generic_polls %>%
  filter(year >= 2014,
         days_until_election <= 31) %>%
  select(pollster, dem, rep, year) %>%
  group_by(year) %>%
  mutate(gen_avg_dem = mean(dem),
         gen_avg_rep = mean(rep)) %>%
  select(year, gen_avg_dem, gen_avg_rep)

# fix the CVAP data to match format of master df
cvap_1 <- cvap_2012_2020 %>%
  filter(year >= 2014) %>%
  select(year, geoid, cvap, moe)

```


```{r}

# CREATING MASTER DF

# creating a master df for district-level modeling, starting with the incumb_dist df as a base
# will only use 2014 to present for model (for now)

data <- incumb_dist %>%
  filter(year >= 2014)

# adding relevant variables from other dfs we have seen over the weeks to the master df

data <- left_join(data, state_econ1, by = c("state" = "State and area", "year" = "Year"))
data <- left_join(data, econ1, by = c("year" = "year"))
data <- left_join(data, expert_rating1, by = c("year" = "year", "state" = "state", "district_num" = "district"))
data <- left_join(data, ratings2018_1, by = c("year" = "year", "state" = "state", "district_num" = "district"))
data <- left_join(data, gp1, by = c("year" = "year"))
data <- left_join(data, cvap_1, by = c("year" = "year", "st_cd_fips" = "geoid"))

# adding a variable for if the district is solid D/R in the form of a yes/no variable
# adding a variable for turnout: coded as the sum of Rep/Dem Votes dividied by CVAP - limitation: this does not account for third party voters
data <- data %>%
  mutate(party_solid = ifelse((avg_rating >= 6 | avg_rating <= 2 | avg_rating_2018 >= 6 | avg_rating_2018 <= 2), "Yes", NA),
         dist_turnout = ((RepVotes + DemVotes)/cvap))

# creating two new dataframes: one for if the district is solid D or R, and another for districts that are not solid D/R
# coded solid as a score if either of the average ratings columns are greater than or equal to 6 (solid R) or less than or equal to 2 (solid D)

solid_dist_data <- data %>%
  filter(party_solid == "Yes") %>%
  select(!c(`7`, `8`, `9`))

battle_dist_data <- data %>%
  filter(is.na(party_solid)) %>%
  select(!c(`7`, `8`, `9`))

```

So far, we have compiled data on election outcomes, generic ballot polls, expert district-level polls, advertising campaigns, and more. When thinking about elections, however, one of the first things that may come to mind are the on-the-ground campaigns. This week we add this "Ground Game" data to our models, specifically with voter turnout at the district level. 

For starters, below we have a basic regression of a district's turnout percentage on the Democratic party's major vote percentage in a district. The turnout percentage is coded as the sum of Democratic votes and Republican votes, divided by the district's citizen voting age population (CVAP).

```{r}
x <- data %>%
  distinct(year, st_cd_fips, DemVotesMajorPercent, dist_turnout)

fit1 <- lm(data = x, DemVotesMajorPercent ~ dist_turnout)
summary(fit1)

```

In this regression, we can see that the intercept is 54.63 percent major vote share for Democrats, representing that when voter turnout is zero, we estimate an average Democratic major vote percentage of 54.63%. When voter turnout is 100%, the Democratic major vote percentage decreases by 5.99 percentage points, to 48.64%. While we do not expect a turnout of 0%, and can only one day hope for a voter turnout of 100%, this model shows us the overall relationship between party vote share and voter turnout. While it is common to think that high voter turnout would benefit Democrats, as they often stress the importance of voting more so than the Republican party, this regression model suggests that higher voter turnout actually *hurts* Democrats. It is worth noting, however, that the R-squared for this model is extremely low, at only 0.0015, essentially stating that there is no relationship between the values of turnout that affect the democratic two party vote percentage. The coefficient on turnout is also not statistically significant, so we don't take much away from this model in terms of prediction power.

```{r}

# add incumbency


```



```{r}

# code from section

cvap_district <- cvap_2012_2020 %>%
  rename(st_cd_fips = geoid) 

# merge with district-level polls
polls_df <- house_polls_long %>%
  filter(year == '2018' | year == '2020')
table(polls_df$year)
# join
cvap_district <- cvap_district %>%
  # filter to relevant years 
  filter(year == '2018' | year == '2020')

polls_cvap_df <- merge(polls_df, cvap_district, by = c('st_cd_fips', 'year'))

# merge with district-level voteshares
dist_pv_df <- incumb_dist
dist_pv_df <- dist_pv_df %>%
  filter(year == '2018' | year == '2020')

polls_cvap_vp_df <- merge(polls_cvap_df, dist_pv_df, by = c('st_cd_fips', 'year'))
table(polls_cvap_vp_df$st_cd_fips)

```




















